Spectral properties of a Wishart-GARCH Matrix

### 2014-12-01 thesis/_region_.tex
1st version that has been handed to Sven. Gaussian Wishart Matrices are normized by 1/T.
This is the version to be used for job searching.

\message{ !name(thesis.tex)}\documentclass{book}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{wrapfig}
\usepackage{amssymb}
\usepackage{amsfonts, amsmath}
\usepackage{mathrsfs}
\usepackage{enumerate}
\usepackage[bookmarks=true]{hyperref}
\usepackage{bookmark}

\input{../physics_common.tex}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}

\title{Mathematical Aspects of the Capital Market}
\author{Xie Xiaolei}
\date{\today}
\begin{document}

\message{ !name(thesis.tex) !offset(1382) }
\cite{NIST:DLMF}:

\message{ !name(thesis.tex) !offset(1837) }

\end{document}

>>>>

\message{ !name(draft4.tex)}\documentclass{article}
\usepackage{amssymb,amsmath,amsthm,bm}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathrsfs}
\usepackage{enumerate}
\usepackage[bookmarks=true]{hyperref}
\usepackage{bookmark}

\input{../physics_common.tex}
\DeclareGraphicsExtensions{.eps, .pdf}

\begin{document}

\message{ !name(draft4.tex) !offset(-3) }

\section{effect of heteroscedasticity in covariance matrix}
Consider a covariance matrix constructed as
$$
C_{ij} = {1 \over T}\sum_{t=1}^T r_{it} r_{jt}
$$
where $1 \leq i \leq N$. Adopt a stochastic volatility model for
$r_{it}$, i.e.
$$
r_{it} = \sigma_{it} \eta_{it}
$$
We can write
\begin{eqnarray*}
C_{ij} &=& {1 \over T}\sum_{t=1}^T \sigma_{it} \eta_{it} \sigma_{jt}
\eta_{jt} \\
\bm{C} &=& {1 \over T} (\bm{\sigma * \eta}) (\bm{\sigma * \eta})'
\end{eqnarray*}
where the matrices $\bm{\sigma}$ and $\bm{\eta}$ have elements
$\sigma_{it}$ and $\eta_{it}$ respectively, and * denotes element-wise
multiplication. Due to the cyclic property of matrix trace, the
moment generating function of $\bm{C}$, $M_C(z)$, relates to $M_D(z)$,
the moment generating function of $\bm D$
\begin{eqnarray*}
  \bm{D} &=& {1 \over T} (\bm{\sigma * \eta})' (\bm{\sigma * \eta})
\end{eqnarray*}
through the equation
\begin{eqnarray*}
  M_C(z) &=& {1 \over q} M_D(z)
\end{eqnarray*}
where $q = N/T$. For later convenience, we make use of $T \times T$
matrices $\bm{\tilde{\sigma}}$ and $\bm{\tilde{\eta}}$, as well as a
projector
$$
\bm{P} = \text{diag}(\underbrace{1, \cdots, 1}_{\text{N 1's}}, 
\underbrace{0, \cdots, 0}_{\text{T-N 0's}})
$$
The first N rows of $\bm{\tilde{\sigma}}$ and $\bm{\tilde{\eta}}$ are
precisely those of $\bm{\sigma}$ and $\bm{\eta}$. This way, we have
\begin{eqnarray*}
\bm D &=& {1 \over T} (\bm{\sigma * \eta})' (\bm{\sigma * \eta}) \\
&=& {1 \over T} (\bm{\tilde{\sigma} * \tilde{\eta}})' \bm P'
\bm P (\bm{\tilde{\sigma} * \tilde{\eta}}) \\
&=& {1 \over T} (\bm{\tilde{\sigma} * \tilde{\eta}})'
\bm P (\bm{\tilde{\sigma} * \tilde{\eta}}) \\
\end{eqnarray*}
Again, by the cyclic property of matrix trace, the spectral properties
of the RHS of the last equation is equivalent to those of
\begin{eqnarray*}
  \bm E &=& {1 \over T} (\bm{\tilde{\sigma} * \tilde{\eta}}) (\bm{\tilde{\sigma}
    * \tilde{\eta}})' \bm P \\
\end{eqnarray*}
In the simple situation where $\sigma_{it} =
\sigma_{jt} = \sigma_t$ for some $\sigma_t$ and for all $i, j, t$,
\begin{eqnarray*}
  \bm{\tilde \sigma * \tilde \eta} &=& \bm{\tilde \eta}
  \begin{pmatrix}
    \sigma_1 &        & \\
        & \ddots & \\
        &        & \sigma_T
  \end{pmatrix} \\
  &=& \bm{\tilde \eta \bar \sigma}
\end{eqnarray*}
Thus
\begin{eqnarray*}
  \bm E &=& {1 \over T}\bm{\tilde \eta \bar \sigma^2 \tilde \eta' P}
\end{eqnarray*}
In the large T limit $\bm{\tilde \eta \bar \sigma^2 \tilde \eta'}$ and $P$ are
freely independent, therefore their S-transforms are multiplicative,
i.e.
\begin{eqnarray*}
  S_E(z) &=& S_{\tilde \eta \bar \sigma^2 \tilde \eta'/T}(z) S_P(z) \\
  &=& S_{\tilde \eta' \tilde \eta /T}(z) S_{\bar \sigma^2}(z) S_P(z)
\end{eqnarray*}
For $S_{\tilde \eta' \tilde \eta /T}(z)$ there have been results when
the elements of $\bm{\tilde \eta}$ are identically distributed with finite
second moment \cite{burda2011} or alternatively, follow $\alpha$-stable
distribution \cite{politi2010}.

So our focus hereafter is on the spectral properties of
$\bar \sigma^2$. The Green's function of $\bar \sigma^2$ can be immediately
written down:
\begin{eqnarray*}
  G_{\bar \sigma^2}(z) &=& {1 \over T} \sum_{t=1}^T \frac{1}{z - \sigma_t^2}
\end{eqnarray*}
If the $\sigma_t$'s are independent, $G_{\bar \sigma^2}(z)$ is deterministic and
equal to $\mean{\frac{1}{z - \sigma_t^2}}$; If, however, the $\sigma_t$'s are
correlated, $G_{\bar \sigma^2}(z)$ is random. In the extreme case where the
$\sigma_t$'s have correlation 1 between any pair, $G_{\bar \sigma^2}(z)$ is a
random variable $\frac{1}{z - \sigma_1^2}$. In the latter case, although
the spectrum of $\bm \bar \sigma^2$ is random, it may still make sense to
know the expectation of the spectrum, which one obtains from the
expectation of the random Green's function. In the rest of this
section we consider a few cases in regard to the distribution of
$\sigma_t$'s as well as their correlations.

\subsection{log-normal volatilities with AR(1) autocorrelation}
In this subsection we consider the case
\begin{eqnarray*}
  \ln \sigma_t &=& \phi\ln \sigma_{t-1} + x_t
\end{eqnarray*}
where $x_t \sim N(0, 1)$. Then, assuming the process was started in
the infinite past, i.e. $t = -\infty$, $\mean{\ln \sigma_t \ln \sigma_s} =
\frac{\phi^{|t-s|}}{1 - \phi^2}$.

\bibliographystyle{unsrt}
\bibliography{econophysics}
\end{document}

\message{ !name(draft4.tex) !offset(-127) }

### 2014-12-01 thesis/price_models2.tex

\S\ref{sec:Garch_model} reviews the GARCH model using the Nordea
15-minute returns as an example. \S\ref{sec:SLV_model} examines the
stochastic log-volatility model and discusses the possibility of
describing the residuals of the log-volotility series using the
Johnson Su distribution. The same Nordea return series is fitted as an
example. \S\ref{sec:XieCalc} derives the unconditional distribution
function of the stochastic log-volotility model and compares the
analytic results to the 30-minute returns of a number of Swedish
stocks.

@@

The picture depicted above is of course overly simplified, and as we discussed in
\S\ref{sec:StylizedFacts}, the distribution of returns is not really
Gaussian. Nevertheless, the assumption of Gaussian distributed returns
underlies such important theories as Black and Scholes theory of
option pricing. Thus, albeit hugely inaccurate, the Gaussian
distribution as a returns model shall not be forgotten. In the next
few sections, we discuss some more realistic returns models.

@@

\section{ARIMA Log-volotility Models}\label{sec:SLV_model}
With the availability of transaction data, estimating conditional
volatility using returns sampled at a higher frequency (realized
volatility) gives superior acuracy and reliability. However, at which
frequency the time series should be sampled in order to give an
unbiased and consistent estimate of the volatility is not a trivial
question. Naively one would argue by the doctrine of statistics that
the often the series is sampled, the better the estimate. This is also
well supported in theory, since the quadratic variation which is the
theoretial variance of the time series (see \cite{Andersen03}, theorem
1 and corollary 1), is approximated as (see \cite{Protter05}, ch.II,
\S6, theorem 22):
\[
\var(p_{t+h} - p_t) = [r, r]_{t+h} - [r, r]_t = \lim_{n \to \infty}
\sum_{T^i \in \pi_n} (p^{T_{i+1}} - p^{T_i})^2
\]
where $p_t$ is the log price process and $r_t = p_t - p_0$. $\pi_n$
is a refining partition of the interval $[t, t+h]$.

@@

To verify that the log-volatility can in general be described by
ARIMA(p, d, q) models, we apply the above procedure to a number of
Swedish stocks whose data have been collected by the author with
home-developed software. In \S\ref{sec:nordea} and \S\ref{sec:volvo}
we model the log-volatilities of Nordea Bank 15-minute returns and
Volvo B 30-minute returns. Both return series span the period
2013/10/10 - 2014/04/04. They are computed using minute-by-minute
average prices, and their variances, i.e. volatilities squared, are
computed as the sum of squared higher-frequency returns, which are in
turn computed using second-by-second prices. Whenever a needed average
price cannot be computed due to the lack of price observations in the
relevent time interval, the value of the previous average price is
assumed.

@@

% \subsection{Ericsson B Log-Volatility}\label{sec:ericsson}
% The probability density function of the residuals $y_t$ that result
% from the MLE estimate is shown in figure
% \ref{fig:nordea_15min_y1_hist} together with its counterpart from the
% initial estimate.
% \subsection{45min Returns}
\section{Unconditional Distribution Functions of ARIMA Log-Volatility
  Models}\label{sec:XieCalc}
In this section we study the unconditional distribution function of
the Stochastic Log-Volatility model specified as equation
\ref{eq:SLV_spec}. As is discussed in \S\ref{sec:SLV_model},
$\ln\sigma_t$ can be described by an ARMA or ARIMA \footnote{ARIMA ---
  Integrated Autoregressive Moving Average} model, possibly with
seasonal components. Here we note that all these models can be
re-written as a moving average model, which is infinite in extent if
autoregressive components are present:
\begin{eqnarray*}
  \ln \sigma_t &=& y_t + \sum_{n=1}^\infty c_n y_{t-n} + \text{Const.}
\end{eqnarray*}
Since the $y_t$ are independent and identically distributed,
\begin{eqnarray*}
  y_t + \sum_{n=1}^\infty c_n y_{t-n}  
\end{eqnarray*}
has Gaussian distribution by the central limit theorem, on condition
that $y_t$ for all $t$ have finite second moment --- this is what we
assume in the rest of this section. It follows from the above equation
that the unconditional distribution of $\ln \sigma_t$ is the same as the
distribution of $\bar{v} + v$ where $v \sim N(0, \sigma)$ and
$\bar{v}$, $\sigma$ are constants. Now we can state that the
unconditional distribution of the returns $r_t$
\begin{eqnarray*}
  r_t &=& \mu + \sigma_t b_t\\
  &=& \mu + \exp\left(
    y_t + \sum_{n=1}^\infty c_n y_{t-n} + \text{Const.}
  \right)
\end{eqnarray*}
 is the same as
\begin{equation}  \label{eq:UnconditionalPdf}
  \begin{aligned}
    r &= \mu + e^{\bar{v} + v} b \\
  \end{aligned}
\end{equation}
where $b \sim N(0, 1)$. For convenience, let $r' = e^v b$

In \S\ref{sec:SLV_Symmetric} we first study the model in the
relatively simple case when $v$ and $b$ are uncorrelated and $\mu
= 0$. If this simplified version proves inadequate, one may resort to
the general model studied in \S\ref{sec:SLV_Asymmetric}.

@@

Following the aforementioned procedure, i.e. computing the initial
estimate by matching the moments and then refining the estimate by
MLE, we obtain the parameter values for a number of return series,
including the Volvo B 30-minute returns described in
\S\ref{sec:SLV_Symmetric}. Table \ref{tab:assets_params} and
\ref{tab:assets_moments} present the obtained parameter values and the
resulting moments, respectively. Figure \ref{fig:Volvo_B_30m_returns},
\ref{fig:Nordea_30m_returns}, and \ref{fig:Ericsson_30m_returns}
compare the empirical distribution functions with the analytic
distribution functions evaluated at these parameters.

@@

In the context of conditional distributions and forecast, this
correlation translates to the correlation between the residual of the
log-volatility, denoted $y_t$ in \S\ref{sec:nordea} and
\S\ref{sec:volvo}, and $b_t$. Now let us consider the forecast
function of an ARIMA model:
\begin{eqnarray*}
  \ln \sigma_t &=& y_t + \sum_{i=1}^P \phi_i \ln \sigma_{t-i} -
  \sum_{i=1}^Q \theta_i y_{t-i}
\end{eqnarray*}
where $\phi_i \leq 1$ if the model involves integration, and $\phi_i <
1$ otherwise. Comparing this equation with equation \ref{eq:r_t}, one
immediately realizes the PDF of $r_t$
\begin{eqnarray*}
  r_t &=& \mu + \sigma_t b_t
\end{eqnarray*}
is given by equation \ref{eq:UncondPDFAsymmetric1} and its moments
given by the equations \ref{eq:AsymmetricMoments} if one makes the
following substitutions:
\begin{eqnarray*}
  \sigma &\to& \text{std}(y_t) \\
  \bar{v} &\to& \sum_{i=1}^P \phi_i \ln \sigma_{t-i} - \sum_{i=1}^Q
  \theta_i y_{t-i} \\
  \psi &\to& \text{corr}(y_t, b_t)
\end{eqnarray*}

For the Nordea series considered in \S\ref{sec:nordea},
$\text{corr}(y_t, b_t)$ is found to be $3.56 \times 10^{-2}$; while
for the Volvo series considered in \S\ref{sec:volvo},
$\text{corr}(y_t, b_t)$ is found to be $-7.3 \times 10^{-3}$. These
values are comparable to those in table \ref{tab:assets_params}, where
$\psi$ for the Nordea series is $1.7234 \times 10^{-2}$ and $-1.5747
\times 10^{-2}$ for the Volvo series. The similarity in these values
provides some evidence about the validity of the model.

>>>>

section \ref{sec:Garch_model} reviews the GARCH model using the Nordea
15-minute returns as an example. Section \ref{sec:SLV_model} examines the
stochastic log-volatility model and discusses the possibility of
describing the residuals of the log-volotility series using the
Johnson Su distribution. The same Nordea return series is fitted as an
example. Section \ref{sec:XieCalc} derives the unconditional distribution
function of the stochastic log-volotility model and compares the
analytic results to the 30-minute returns of a number of Swedish
stocks.

@@

The picture depicted above is of course overly simplified, and as we discussed in
section \ref{sec:StylizedFacts}, the distribution of returns is not really
Gaussian. Nevertheless, the assumption of Gaussian distributed returns
underlies such important theories as Black and Scholes theory of
option pricing. Thus, albeit hugely inaccurate, the Gaussian
distribution as a returns model shall not be forgotten. In the next
few sections, we discuss some more realistic returns models.

@@

\section{ARIMA Log-volotility Models}\label{sec:SLV_model}
With the availability of transaction data, estimating conditional
volatility using returns sampled at a higher frequency (realized
volatility) gives superior acuracy and reliability. However, at which
frequency the time series should be sampled in order to give an
unbiased and consistent estimate of the volatility is not a trivial
question. Naively one would argue by the doctrine of statistics that
the often the series is sampled, the better the estimate. This is also
well supported in theory, since the quadratic variation which is the
theoretial variance of the time series (see \cite{Andersen03}, theorem
1 and corollary 1), is approximated as (see \cite{Protter05}, ch.II,
section 6, theorem 22):
\[
\var(p_{t+h} - p_t) = [r, r]_{t+h} - [r, r]_t = \lim_{n \to \infty}
\sum_{T^i \in \pi_n} (p^{T_{i+1}} - p^{T_i})^2
\]
where $p_t$ is the log price process and $r_t = p_t - p_0$. $\pi_n$
is a refining partition of the interval $[t, t+h]$.

@@

To verify that the log-volatility can in general be described by
ARIMA(p, d, q) models, we apply the above procedure to a number of
Swedish stocks whose data have been collected by the author with
home-developed software. In section \ref{sec:nordea} and section \ref{sec:volvo}
we model the log-volatilities of Nordea Bank 15-minute returns and
Volvo B 30-minute returns. Both return series span the period
2013/10/10 - 2014/04/04. They are computed using minute-by-minute
average prices, and their variances, i.e. volatilities squared, are
computed as the sum of squared higher-frequency returns, which are in
turn computed using second-by-second prices. Whenever a needed average
price cannot be computed due to the lack of price observations in the
relevent time interval, the value of the previous average price is
assumed.

@@

% \subsection{Ericsson B Log-Volatility}\label{sec:ericsson}
% The probability density function of the residuals $y_t$ that result
% from the MLE estimate is shown in figure
% \ref{fig:nordea_15min_y1_hist} together with its counterpart from the
% initial estimate.
% \subsection{45min Returns}
\section{Unconditional Distribution Functions of ARIMA Log-Volatility
  Models}\label{sec:XieCalc}
In this section we study the unconditional distribution function of
the Stochastic Log-Volatility model specified as equation
\ref{eq:SLV_spec}. As is discussed in section \ref{sec:SLV_model},
$\ln\sigma_t$ can be described by an ARMA or ARIMA \footnote{ARIMA ---
  Integrated Autoregressive Moving Average} model, possibly with
seasonal components. Here we note that all these models can be
re-written as a moving average model, which is infinite in extent if
autoregressive components are present:
\begin{eqnarray*}
  \ln \sigma_t &=& y_t + \sum_{n=1}^\infty c_n y_{t-n} + \text{Const.}
\end{eqnarray*}
Since the $y_t$ are independent and identically distributed,
\begin{eqnarray*}
  y_t + \sum_{n=1}^\infty c_n y_{t-n}  
\end{eqnarray*}
has Gaussian distribution by the central limit theorem, on condition
that $y_t$ for all $t$ have finite second moment --- this is what we
assume in the rest of this section. It follows from the above equation
that the unconditional distribution of $\ln \sigma_t$ is the same as the
distribution of $\bar{v} + v$ where $v \sim N(0, \sigma)$ and
$\bar{v}$, $\sigma$ are constants. Now we can state that the
unconditional distribution of the returns $r_t$
\begin{eqnarray*}
  r_t &=& \mu + \sigma_t b_t\\
  &=& \mu + \exp\left(
    y_t + \sum_{n=1}^\infty c_n y_{t-n} + \text{Const.}
  \right)
\end{eqnarray*}
 is the same as
\begin{equation}  \label{eq:UnconditionalPdf}
  \begin{aligned}
    r &= \mu + e^{\bar{v} + v} b \\
  \end{aligned}
\end{equation}
where $b \sim N(0, 1)$. For convenience, let $r' = e^v b$

In section \ref{sec:SLV_Symmetric} we first study the model in the
relatively simple case when $v$ and $b$ are uncorrelated and $\mu
= 0$. If this simplified version proves inadequate, one may resort to
the general model studied in section \ref{sec:SLV_Asymmetric}.

@@

Following the aforementioned procedure, i.e. computing the initial
estimate by matching the moments and then refining the estimate by
MLE, we obtain the parameter values for a number of return series,
including the Volvo B 30-minute returns described in
section \ref{sec:SLV_Symmetric}. Table \ref{tab:assets_params} and
\ref{tab:assets_moments} present the obtained parameter values and the
resulting moments, respectively. Figure \ref{fig:Volvo_B_30m_returns},
\ref{fig:Nordea_30m_returns}, and \ref{fig:Ericsson_30m_returns}
compare the empirical distribution functions with the analytic
distribution functions evaluated at these parameters.

@@

In the context of conditional distributions and forecast, this
correlation translates to the correlation between the residual of the
log-volatility, denoted $y_t$ in section \ref{sec:nordea} and
section \ref{sec:volvo}, and $b_t$. Now let us consider the forecast
function of an ARIMA model:
\begin{eqnarray*}
  \ln \sigma_t &=& y_t + \sum_{i=1}^P \phi_i \ln \sigma_{t-i} -
  \sum_{i=1}^Q \theta_i y_{t-i}
\end{eqnarray*}
where $\phi_i \leq 1$ if the model involves integration, and $\phi_i <
1$ otherwise. Comparing this equation with equation \ref{eq:r_t}, one
immediately realizes the PDF of $r_t$
\begin{eqnarray*}
  r_t &=& \mu + \sigma_t b_t
\end{eqnarray*}
is given by equation \ref{eq:UncondPDFAsymmetric1} and its moments
given by the equations \ref{eq:AsymmetricMoments} if one makes the
following substitutions:
\begin{eqnarray*}
  \sigma &\to& \text{std}(y_t) \\
  \bar{v} &\to& \sum_{i=1}^P \phi_i \ln \sigma_{t-i} - \sum_{i=1}^Q
  \theta_i y_{t-i} \\
  \psi &\to& \text{corr}(y_t, b_t)
\end{eqnarray*}

For the Nordea series considered in section \ref{sec:nordea},
$\text{corr}(y_t, b_t)$ is found to be $3.56 \times 10^{-2}$; while
for the Volvo series considered in section \ref{sec:volvo},
$\text{corr}(y_t, b_t)$ is found to be $-7.3 \times 10^{-3}$. These
values are comparable to those in table \ref{tab:assets_params}, where
$\psi$ for the Nordea series is $1.7234 \times 10^{-2}$ and $-1.5747
\times 10^{-2}$ for the Volvo series. The similarity in these values
provides some evidence about the validity of the model.
